---
title: "Cities of the World"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(mosaic)
library(mosaicData)
library(ggplot2)
library(USCOTS2017DC)
library(ggformula)
library(DataComputing)
knitr::opts_chunk$set(echo = FALSE)
```

## Cities of the World

The data table `WorldCities` identifies `r nrow(WorldCities)` cities around the world that have large populations or are large for their region. Here are a few randomly selected rows from the table:

```{r}
set.seed(101)
WorldCities %>%
  mutate(continent = gsub("/.*$", "", region)) %>%
  group_by(continent) %>%
  filter(n() > 100, row_number(desc(population)) > 10) %>%
  sample_n(size = 3) %>%
  ungroup() %>%
  sample_n(size = 6) %>%
  select(-continent)
```

### Exercise

Write a wrangling statement that finds the total population of all the cities in `WorldCities`. Compare the result to the population of the world, which you can find on the internet.

```{r cities0, exercise = TRUE}
WorldCities %>%
  ___(total = ___(population))
```

## Starting big

How many cities with a population larger than 100,000? Larger than 1,000,000? Larger than 10,000,000?

```{r cities1, exercise = TRUE}
WorldCities %>%

```


## City locations

Make a scatterplot of the `latitude` and `longitude` of cities larger than 100K.

```{r cities2, exercise = TRUE}
WorldCities %>%
  filter(___) %>%
  gf_point(___)
```

Once you have the plot made, modify it a bit:

* There are so many cities that the dots lie on top of one another. To get a better display of how many cities are in each location, use transparency, called `alpha`, to handle overplotting.  Alpha can run from zero to one: zero is completely transparent (a.k.a. invisible); one is completely opaque.

* Find a way to work the city `population` into the graphic. Some possibilities: color, size, alpha. Try several possibilities until you get a plot that is informative.

* When you have your plotting commands complete, use those commands to make another graphic, but add this expression to govern the `size` attribute: `%>% gf_refine(scale_size_area())`.  This will make the *area* of the dot proportional to the value of the variable mapped to it.  Without `scale_size_area()`, the *diameter* of the dot is proportional to the variable.  Explain which scale, area or diameter, you think is most informative.  (Include both graphics in your Rmd file along with your explanation.)



## Biggest city

The following will output a data table with the single biggest city in the world.
```{r cities3, exercise = TRUE}
BiggestByCountry <-
  WorldCities %>%
  filter( rank(desc(population)) == 1 )
BiggestByCountry
```

Modify the statement so that the output contains the biggest city in each country. Arrange this from biggest population to smallest.

## Plotting the biggest

Assemble the commands from previous chunks to make a plot of the locations of all cities. Add a layer to the plot showing the location and size of the biggest city in each country. Show
that city in red.

```{r cities4, exercise = TRUE}
BiggestByCountry <- _copy_your_statements_here_
WorldCities %>%
  gf_point(____) %>%
  gf_point(____, data = BiggestByCountry) %>%
  gf_text(____)
```

Modify the graph you made to add another layer with with `gf_text()` to show the name of the biggest city. Map `name` to the `label` aesthetic in `gf_text()`. You can also use the `size` aesthetic to control the label size.


## Biggest in each country.

Create a table that shows the biggest city in each country but just for cities above 5M in population. Display as output all the cases but just these variables: city name, country, and population.

```{r cities5, exercise = TRUE}

```
## Urban population

Some countries are more urbanized than others. Although `WorldCities` doesn't give the *total population* of each country, those data are contained in the `pop` variable in the `CountryData` table.

Your job is to create a display showing the fraction of the population of each country who live in cities larger than 50,000.

To do this, you will need to *join* the `WorldCities` and `CountryData` tables. Regretably, `WorldCities` and `CountryData` do not use the same code to identify countries. A translation between them is contained in the `CountryCodes` table.

Create an informative display of the fraction of the total population living in cities larger than 50,000 people.

Here is some scratch space for you to try things out ...

```{r cities6, exercise = TRUE, exercise.cap = "Scratch space"}

```

Assemble the all the commands to join the tables and to make your display here.

```{r cities7, exercise = TRUE, exercise.cap = "Assembled commands"}

```




